Lessons learned from bugs in models of human history

### 2020-05-22 paper.tex
Some edits after comments

Fortunately, the effects of this error are subtle.
Population sizes and structure since the time of
the earliest split are unaffected, so differences in expected $F_{ST}$ are negligible between
the intended and incorrect models. However, the ancient structure distorts the distribution
of early TMRCAs. The extraneous ancient population increases the long-term effective
population size, resulting in roughly $4\%$ excess heterozygosity in
contemporary populations compared to the intended model
(Fig.~\ref{fig:ooa_stats}C-F).

@@

In another publication using this model~\citep{martin2017human},
a separate error was introduced: the model itself was defined as suggested
in the documentation (using updated parameters
from~\citet{gravel2011demographic}) and inspected
using the \msprime\ debugging tools.
Despite these initial checks being made, the simulation
was performed without passing the parameter defining demographic events,
so that the three populations
never merged and remained separated with low levels of migration (Fig~\ref{fig:prs}A),
leading to a vast overestimate of the divergence across human populations.
While the correct model predicts a mean $F_{ST}$ of
$0.05 - 0.10$ across the three population, the simulated model generated $F_{ST}$
ranging between $0.3 - 0.6$, depending on the populations considered.
Overall diversity was also strongly affected: heterozygosity was more than
doubled in African populations and reduced by more than half in Eurasian populations
relative to the intended model.

@@

Difficulties in transferability remain in the intended model (Fig.~\ref{fig:prs}), although
risk prediction in each population is significantly improved (compare to Fig.~5 in~\citet{martin2017human}).
The intended simulations indicate that the accuracy of genetic risk scores is still
substantially reduced in understudied populations (Fig.~\ref{fig:prs}E-G),
supporting one of the main conclusions of~\citet{martin2017human}.
However, the reduction is much less pronounced than reported.
In particular, we do not observe large differences in mean predicted risk 
across populations (Fig.~\ref{fig:prs}C,D), as presented in~\citet{martin2017human}.
At least for the neutral polygenic architectures considered here, genetic drift alone
does not appear to induce large directional biases in mean predicted risk.

This error was easy to make, and could have been prevented by better API design.
The original \msprime\ API requires the user to pass
three separate parameters to specify a demographic model. For example,
\begin{lstlisting}[frame=single]
dbg = msprime.DemographyDebugger(
  population_configurations=population_configurations,
  demographic_events=demographic_events,
  migration_matrix=migration_matrix)
dbg.print_history()
ts = msprime.simulate(
  population_configurations=population_configurations,
  demographic_events=demographic_events, # Missing in Martin et al. 2017!
  migration_matrix=migration_matrix)
\end{lstlisting}
Note that the same three parameters must be passed to both the debugger
and the simulate function.
To help prevent such errors,
\msprime~ver.~1.0 will introduce a Demography class. The above
snippet would then be rewritten as
\begin{lstlisting}[frame=single]
demography = msprime.Demography(
  population_configurations=population_configurations,
  demographic_events=demographic_events,
  migration_matrix=migration_matrix)
dbg = demography.debug()
dbg.print_history()
ts = msprime.simulate(demography=demography)
\end{lstlisting}
This simple change to the interface makes it much less likely
that different models are passed to the debugger and simulator,
reducing the potential for error.

@@

\begin{figure}[ht]
\begin{center}
\makebox[\textwidth][c]{\includegraphics{figures/ooa_expected_stats.pdf}}
\caption{\textbf{Expected diversity statistics under the \citet{gutenkunst2009inferring} model}.
    \textbf{(A)} The correctly implemented model. Dashed arrows depict continuous migration.
    \textbf{(B)} The incorrectly implemented model from the \msprime\ tutorial, with migration continuing
    into the past beyond the mass migration event with proportion $1$ from the ancestral population
    to the bottleneck population.
    \textbf{(C)} Marginal allele frequency spectra under the two models. Heterozygosity in the incorrect model
    is inflated by $~3.5\%$, though the general shape of the distributions are qualitatively similar.
    \textbf{(D)} Similarly, the increased heterozygosity leads to excess $D^2$, though the LD-decay is
    qualitatively similar between models.
    \textbf{(E-G)} True size history for each population plotted against the expected size history from
    the expected inverse coalescence rates.
}
\label{fig:ooa_stats}
\end{center}
\end{figure}

\begin{figure}[ht]
\begin{center}
\makebox[\textwidth][c]{\includegraphics{figures/prs_fig.pdf}}
\caption{\textbf{The transferability of PRS under neutrality}.
    \textbf{(A)} In~\citet{martin2017human}, the simulated demographic model did not apply demographic
    events in the past, so continental populations were simulated as isolated with low levels of migration
    for all time. The intended model is shown in Figure \ref{fig:ooa_stats}A.
    \textbf{(B-G)} We repeated the simulation experiment in~\cite{martin2017human} using the correct
    demographic model.
    While risk prediction in the African and East Asian population is reduced compared to the studied
    European population, the reduction in prediction accuracy is not as hopeless as originally reported.
    Notably, under a neutral polygenic architecture, we do not expect significant bias in inferred PRS
    in any of the populations (C, D).
    Correlations were computed over 100 simulation replicates.
    For direct comparison to the original study, see Figure~5 in~\citet{martin2017human}.
}
\label{fig:prs}
\end{center}
\end{figure}

>>>>

Fortunately, the effects of this error are subtle.
Population sizes and structure since the time of
the earliest split are unaffected, so differences in expected $F_{ST}$ are negligible between
the correct and incorrect models. However, the ancient structure distorts the distribution
of early TMRCAs (Fig.~\ref{fig:ooa_stats}E-G).
The extraneous ancient population increases the long-term effective
population size, resulting in roughly $4\%$ excess heterozygosity in
contemporary populations compared to the intended model, but the overall effects
on patterns of diversity are minimal (Fig.~\ref{fig:ooa_stats}C,D).

@@

In another publication using this model~\citep{martin2017human},
a separate error was introduced: the model itself was defined as suggested
in the documentation (using updated parameters
from~\citet{gravel2011demographic}) and inspected
using the \msprime\ debugging tools.
Despite these initial checks being made, the simulation
was performed without passing the parameter defining demographic events,
so that the three populations
never merged and remained separated with low levels of migration (Fig~\ref{fig:prs}A),
leading to a vast overestimate of the divergence across human populations.
While the correct model predicts a mean $F_{ST}$ of
$0.05 - 0.10$ across the three population, the simulated model generated $F_{ST}$
ranging between $0.3 - 0.6$, depending on the populations considered.
Overall diversity was also strongly affected: heterozygosity was more than
doubled in African populations and reduced by more than half in Eurasian populations
relative to the correct model.

@@

Difficulties in transferability remain in the corrected model (Fig.~\ref{fig:prs}), although
risk prediction in each population is significantly improved (compare to Fig.~5 in~\citet{martin2017human}).
Simulations under the correct model indicate that the accuracy of genetic risk scores is still
substantially reduced in understudied populations (Fig.~\ref{fig:prs}E-G),
supporting one of the main conclusions of~\citet{martin2017human}.
However, the reduction is much less pronounced than reported.
In particular, we do not observe large differences in mean predicted risk 
across populations (Fig.~\ref{fig:prs}C,D), as presented in~\citet{martin2017human}.
At least for the neutral polygenic architectures and continental-scale demographic model
considered here, genetic drift alone does not appear to induce large directional biases
in mean predicted risk.

This error was easy to make, and could have been prevented by better API design.
The original \msprime\ API requires the user to pass
three separate parameters to specify a demographic model. For example,
\begin{lstlisting}[frame=single]
dbg = msprime.DemographyDebugger(
  population_configurations=population_configurations,
  demographic_events=demographic_events,
  migration_matrix=migration_matrix)
dbg.print_history()
ts = msprime.simulate(
  population_configurations=population_configurations,
  demographic_events=demographic_events, # Missing in Martin et al. 2017
  migration_matrix=migration_matrix)
\end{lstlisting}
Note that the same three parameters must be passed to both the debugger
and the simulate function.
To help prevent such errors,
\msprime~ver.~1.0 will introduce a Demography class. The above
snippet would then be rewritten as
\begin{lstlisting}[frame=single]
demography = msprime.Demography(
  population_configurations=population_configurations,
  demographic_events=demographic_events,
  migration_matrix=migration_matrix)
dbg = demography.debug()
dbg.print_history()
ts = msprime.simulate(demography=demography)
\end{lstlisting}
This simple change to the interface makes it much less likely
that different models are passed to the debugger and simulator,
reducing the potential for error.

@@

\begin{figure}[ht]
\begin{center}
\makebox[\textwidth][c]{\includegraphics{figures/ooa_expected_stats.pdf}}
\caption{\textbf{Expected diversity statistics under the \citet{gutenkunst2009inferring} model}.
    \textbf{(A)} The correctly implemented model. Dashed arrows depict continuous migration.
    \textbf{(B)} The incorrectly implemented model from the \msprime\ tutorial, with migration continuing
    into the past beyond the mass migration event with proportion $1$ from the ancestral population
    to the bottleneck population.
    \textbf{(C)} Marginal allele frequency spectra under the two models. Heterozygosity in the incorrect model
    is inflated by $\sim3.5\%$, though the general shape of the distributions are qualitatively similar.
    \textbf{(D)} Similarly, the increased heterozygosity leads to excess $D^2$, though the LD-decay is
    qualitatively similar between models.
    \textbf{(E-G)} True size history for each population plotted against the expected size history computed
    from the inverse coalescence rates under each model.
}
\label{fig:ooa_stats}
\end{center}
\end{figure}

\begin{figure}[ht]
\begin{center}
\makebox[\textwidth][c]{\includegraphics{figures/prs_fig.pdf}}
\caption{\textbf{The transferability of PRS under neutrality}.
    \textbf{(A)} In~\citet{martin2017human}, the simulated demographic model did not apply demographic
    events in the past, so continental populations were simulated as isolated with low levels of migration
    for all time. The correct model has historical events as shown in Fig.~\ref{fig:ooa_stats}A.
    \textbf{(B-G)} We repeated the simulation experiment in~\cite{martin2017human} using the correct
    demographic model.
    GWAS summary statistics were computed from 10,000 case and control subjects in the European
    population, and the distribution of inferred PRS were compared across the three simulated populations.
    Unlike the original study, we do not observe large differences in mean inferred PRS across
    the three populations (C, D), and while risk prediction in the African and East Asian population is
    still reduced compared to the European population, the reduction in prediction accuracy is not as
    large as reported in the original study (E-G).
    Correlations between true and inferred polygenic scores were computed over 100 simulation replicates.
    For direct comparison to the original study, see Figure~5 in~\citet{martin2017human}.
}
\label{fig:prs}
\end{center}
\end{figure}
